---
layout: module
title: Syntax
date: 2018-01-01 00:00:02
category: module
links:
  script: syntax.R
output:
  md_document:
    variant: markdown_mmd
    preserve_yaml: true
---

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> R Workshop
## <FILE> syntax.R 
## <INIT> 15 January 2018
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################

## clear memory
rm(list = ls())

```

It's hard to know where to start when teaching a new programming
language. This page is meant to give some background about R that
hopefully  
1. explains a little about how it is put together, and  
2. puts it in context with other programming languages you might know.

That said, revisiting this page after after working through the other
modules might be useful.

# R: language + environment
```{r, include = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Language and environment
## ---------------------------------------------------------

## run me first
print('Hello, World!')

```

R is a port of the [S
language](https://en.wikipedia.org/wiki/S_(programming_language)),
which was developed at Bell Labs. As a [GNU
project](https://www.gnu.org), R is open source and free ([as in
freedom](https://en.wikipedia.org/wiki/Gratis_versus_libre)) to use
and distribute. It can be installed and used on most major operating
systems.

R is best thought of as an integrated [language and
environment](https://www.r-project.org/about.html) that was designed
with statistical computing and data analysis in mind. To that end, its
structure represents a compromise between a code base optimized for
mathematical procedures and one with high-level functionality that can
be used interactively (unlike compiled code). In other words, it's a
great tool for working interactively with quantitative data.

R is probably best known
for its graphing capabilities, but it has [continued to grow in
popularity among data
scientists](http://blog.revolutionanalytics.com/2018/01/tiobe-2017.html),[^f1]
who are increasingly [extending R's functionality through user-contributed
packages](http://blog.revolutionanalytics.com/2017/01/cran-10000.html).[^f2]
We will use a number of packages in this workshop.

## Integrated development environment (IDE) for R

### RStudio

[RStudio](https://www.rstudio.com) does most everything R-related well
and with little fuss, so it's a great all-around program for using R. 
We will use it in this workshop.

> #### Quick exercise
> If you haven't already, open up RStudio and poke around. First, try
> entering an equation in the console (like `1 + 1`). Next, open the
> script associated with this module and run the first line.

# Assignment
```{r, include = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Assignment
## ---------------------------------------------------------

```
R thinks of things as objects. Objects are like boxes in which we can
put things: data, functions, and even other objects.

Before discussing data types and structures, the first lesson in R is
how to assign values to objects. In R ([for quirky
reasons](http://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html)),
the primary means of assignment is the arrow, `<-`, which is a less
than symbol, `<`, followed by a hyphen, `-`.

```{r}
## assign value to object x using <-
x <- 1

## show
x

```

You can also assign using a single equals sign, `=`:
```{r}
## assign value to object y using =
y = 'a'

## show
y

```

Keep in mind, however, that since `=` sometimes has
other meanings in R (it's how functions set argument options), it may
be clearer to use `<-`.

> #### Quick exercise
> Using the arrow, assign the output of `1 + 1` to `x`. Next subtract 1
> from `x` and reassign the result to `x`.

# Data types and structures
```{r, include = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Data types and structures
## ---------------------------------------------------------
```

## Types  
```{r, include = FALSE, purl = TRUE}
## ---------------------------
## types
## ---------------------------
```

There are three primary data types in R that you will
regularly use:  
- `logical`  
- `numeric` (`integer` & `double`)  
- `character` 

### Logical

Logical vectors can be `TRUE`, `FALSE`, or `NA`. They can be assigned
to objects or returned by logical operators (*e.g.*, `==`, `!=`, `<`,
`>`, etc), which makes them useful for control flow in loops and
functions.

**NB** In R, you can shorten `TRUE` to `T` and `FALSE` to `F`, but
both the short and long versions must be capitalized.

```{r, include = FALSE, purl = TRUE}
## ------------
## logical
## ------------

```
```{r}
## assignment
x <- TRUE
x

## ! == NOT
!x

## check
is.logical(x)

## evaluate
1 + 1 == 2

```

### Numeric: Integer and Double

Numeric values can be both **integers** and double precision floating
point values, or just **doubles**. R automatically converts between the two data
types for you, so knowing the difference between the two isn't really
important for most analyses. 

If you want to use an integer, place a capital `L` after the number
like `1L`. If a number is stored as an integer, some R output will
place an `L` behind the digits to let you know that. Mostly, R
defaults to using doubles, but if you see a number with an `L` behind
it, know that it's still a number.


```{r, include = FALSE, purl = TRUE}
## ------------
## numeric
## ------------

```
```{r}
## use 'L' after digit to store as integer
x <- 1L
is.integer(x)

## R stores as double by default
y <- 1
is.double(y)

## both are numeric
is.numeric(x)
is.numeric(y)

```

### Character

Character values are stored as strings, which means you need to place
either single `'` or double `"` quotes around them. Numeric values can
also be stored as strings (sometimes useful if you must store leading
zeroes), but they have to be converted back to numbers before you can
perform numeric operations on them (like adding or subtracting) or use
them in a statistical model.

```{r, include = FALSE, purl = TRUE}
## ------------
## character
## ------------

```
```{r}
## store a string using quotation marks
x <- 'The quick brown fox jumps over the lazy dog.'
x

## store a number with leading zeros
x <- '00001'
x

```

> #### Quick exercise
> Try to add a string digit to a numeric value. What happens? Can you
> convert the string version on the fly so that the equation works?
> (HINT: in R, you can change a vector type using `as.<type>()`,
> where `<type>` is the name of what you want.)

## Structures
```{r, include = FALSE, purl = TRUE}
## ---------------------------
## structures
## ---------------------------
```
Building on these data types, R relies on four primary data structures:

- `vector`
- `matrix`[^f6]
- `list`
- `dataframe`

### Vector

A vector in R is just a collection of the data types discussed. In
fact, a single value is a vector of one. Vectors do not have
dimensions (`dim()`), but do have `length()`, which is good to
remember when inspecting your data or writing loops and functions. 

You combine multiple values using the concatenate, `c()`, function. We
will use `c()` a lot.

```{r, include = FALSE, purl = TRUE}
## ------------
## vector
## ------------

```  
```{r}
## create vector
x <- 1

## check
is.vector(x)

## add to vector (can do so recursively meaning old x can help make new x)
x <- c(x, 5, 8)
x

## no dim...
dim(x)

## ...but length
length(x)

```

You can access the elements of a vector using brackets, `[]`, after
the object name. If you think of each element in the vector as having
an address, that is, a way to access it specifically, then its address
is its position number in the vector. This position number is called
its index, and in R, the index always starts with `1`.

In our current vector, we have three items, ```r x[1]```, ```r x[2]```, 
and ```r x[3]```, which in turn have indices of `1`, `2`,
and `3`. To access ```r x[2]``` specifically, we can call it using the
brackets and its index: `x[2]`.

```{r}
## get the second element
x[2]

```

> #### Quick exercise 
> Since you know how to access a specific element
> in a vector and how to assign new values, try to change the 3rd
> element of the `x` vector to 4.

All values in a vector must be of the same type. If you concatenate
values of different data types, R will automatically promote all
values to least ambiguous type. We can check this with `class()`.

```{r}
## check class of x
class(x)

## add character
x <- c(x, 'a')
x

## check class
class(x)

```

### Matrix
```{r, include = FALSE, purl = TRUE}
## ------------
## matrix/array
## ------------

```  

A matrix is a 2D arrangement of data types. Instead of length, it has
dimensions. Like vectors, all data elements must be of the same type.

```{r}
## create 3 x 3 matrix that is the sequence of numbers between 1 and 9
x <- matrix(1:9, nrow = 3, ncol = 3)
x

## ...fill by row this time
y <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
y

## a matrix has dimension
dim(x)

```

Use `nrow()` and `ncol()` to get the number of rows and columns,
respectively.


```{r}
## # of rows
nrow(x)

## # of columns
ncol(x)

```

Like a vector, you can access parts of a matrix. Since it has two
dimensions, use a comma in the bracket to separate row indices from
column indices.

When using brackets with objects that have two dimensions, a
good rule of thumb is to add your comma first: `x[ , ]`. Numbers or
objects you put between the first bracket and the comma will affect
the rows; numbers between the comma and the closing bracket will
affect the columns.

If you don't put anything in either of those spaces (a blank space
doesn't count), R will assume you want all rows or columns, depending
on which side of the comma is blank.

```{r}
## show the values in the first row
x[1, ]

## show the values in the third column
x[ ,3]

## this is the same as just calling x by itself
x[ , ]

```

> #### Quick exercise
> Return the middle value of the `x` matrix. Next assign the middle
> value the character value 'a'. What happens to the rest of the
> values in the matrix?


### List
```{r, include = FALSE, purl = TRUE}
## ------------
## list
## ------------

```  

Lists are a catch all objects that can hold an assortment of other
objects of different data types. They can be flat, meaning that all
values are at the same level, or nested, with lists holding other
lists.

```{r}
## create single-level list
x <- list(1, 'a', TRUE)

## show
x

## check
is.list(x)

## create blank list
y <- list()

## add to first list, creating nested list
z <- list(x, y)

## show
z

```

You access items in lists like you do vectors and matrices. You may,
however, need to use double brackets, `[[]]`, and multiple pairs,
`[[]][[]]`, to reach the item you need.

```{r}
## the first item in list z is list x
z[[1]]

## to get to 'a' in list x, need to add more brackets
z[[1]][[2]]

```

### Data frame
```{r, include = FALSE, purl = TRUE}
## ------------
## data frame
## ------------

```
Data frames are really just an organized collection of lists / vectors that
are the same length. That quick description, however, belies the
importance of data frames: you will use them all the time in your
data work.

Most of the time, you will be reading in data frames, but you can also
create them.

```{r}
## create data frame where col_* are the column (variable) names
df <- data.frame(col_a = c(1,2,3),
                 col_b = c(4,5,6),
                 col_c = c(7,8,9))

## show
df

## check
is.data.frame(df)
    
```

Like matrices, data frames have a `dim()` and the number of rows and
columns can be recovered using `nrow()` and `ncol()`. The column
names, which are needed when estimating models and making graphics,
are accessed using `names()`.

```{r}
## get column names
names(df)

```

To access a column, you need to give R the data frame's name followed
by a `$` and then the variable name.


```{r}
## get col_a
df$col_a

```

You can also use the `df[['<var name>']]` construction, which comes in
handy in loops and functions.

```{r}
## get col_a (note the quotation marks this time)
df[['col_a']]

```
> #### Quick exercise 
> Create two or three equal length vectors. Next, combine to create a
> data frame. Finally, change one value in the data frame (HINT: think
> about how you changed vector and matrix values before).

# Packages

User-submitted packages are a huge part of what makes R great. Most of
your scripts will make use of one or more packages.

## Installation
### CRAN

As you've seen on the <a href = '{{ site.baseurl }}/start/'>getting
started</a> page, packages can be installed from the official CRAN
repository using:

```{r, eval = FALSE, purl = FALSE}
install.packages('<package name>')
```

The default option installs all dependencies (other packages that the
package you want may rely on to work properly). By default, R will
check how you installed R and download the right operating system file
type.

> #### Quick exercise 
> Install the
> [`survey`](https://CRAN.R-project.org/package=survey) package,
> which we will use in a later module. Don't forget to use single or
> double quotation marks around the package name.

### GitHub

Recently, people have begun sharing the source code for their R
packages on [GitHub](https://github.com). If you want to download a
package on GitHub, either because it isn't hosted on CRAN or because
you want the newest development version, you can use the `devtools`
package to get it (you will need [`git`](https://git-scm.com) on your
system, too):  

```{r, eval = FALSE, purl = FALSE}
library(devtools)
install_github('<github handle>/<repo name>')
```
## Loading package libraries

Package libraries[^f7] can loaded in a number of ways, but the easiest
it to write:

```{r, eval = FALSE, purl = FALSE}
library('<library name>')
```

where `'<library name>'` is the name of the package/library. You will
need to load these before you can use their functions in your
scripts. Typically, they are placed at the top of the script file.

> #### Quick exercise 
> Load the `tidyverse` package, which you should have already
> installed. This will be a good test of the installation since we
> will use tidyverse libraries throughout the rest of the workshop.

# Help
```{r, echo = FALSE, purl = TRUE}
## ---------------------------------------------------------
## Help
## ---------------------------------------------------------

```

Even I don't have every R function and nuance memorized. With all the
user-written packages, it would be difficult to keep up if I tried!
When stuck, there are a few ways to get help.

## Help files

In the console, typing a function name immediately after a question
mark will bring up that function's help file:

```{r, eval = FALSE}
## get help file for function
?sum
```

Two question marks will search for the command name in CRAN packages:

```{r, eval = FALSE}
## search for function in CRAN
??sum
```

## Google it!

Google is a coder's best friend. If you are having a problem, odds are
a 1,000 other people have too and at least one of them has been brave
(or foolhardy!) enough to ask about it in a forum like
[StackOverflow](https://stackoverflow.com),
[CrossValidated](https://stackoverflow.com), or [R-help mailing
list](https://stat.ethz.ch/mailman/listinfo/r-help). Google it!

# Miscellaneous notes about R

## Compared to other statistical languages

Like all computing languages, R has its own structure and quirks. The
idiomatic R approach to data analysis can be especially challenging at
first for those who come to R from other common statistical packages
or scripting languages, like
[SPSS](https://www.ibm.com/products/spss-statistics),
[Python](https://www.python.org), and
[Stata](https://www.stata.com).[^f3] 

I came to R after learning Stata first, which I think is common for
many researchers trained in econometric methods. For me and others
who've made the same Stata-to-R transition, I think the root of many
problems is the fundamental difference between how Stata and R
operate. Whereas Stata is more of
a
[procedural language](https://en.wikipedia.org/wiki/Procedural_programming) in
which commands **do** things in an environment (your data), R is
more
[object-oriented](https://en.wikipedia.org/wiki/Object-oriented_programming) in
that data and functions are **stored** in variables or objects and
await instructions that pertain to them.[^f4]

As pointed out by my friend and colleague [Richard
Blissett](https://www.shu.edu/profiles/RichardBlissett.cfm), users can
see this difference in the command/function names in each
language. Stata commands tend to be verbs: `summarize`, `tabulate`,
and `regress`; on the other hand, R functions are often nouns:
`summary`, `table`, and `lm` (for linear model). And so, common
problems in the R to Stata switch such as  
- *I ran a model and didn't get any output...*  
- *How do I create local/global macros in R?*  
- *Which of these data objects is the actual data?*  
- *Why isn't R **doing** anything?*  

may be due to misunderstanding this difference.[^f5]

Like learning a new spoken language, constantly translating between
your native tongue and the new language will only get you so far. To
that end, I encourage native-Stata users to try to approach R without
Stata procedures in mind (easier said than done, I know). That said,
this document that shows the same analysis done in [Stata and R
side-by-side](http://rslblissett.com/wp-content/uploads/2016/09/sidebyside_130826.pdf)
may be useful in the initial transition.

## Other options for running R

There are many other ways besides RStudio to run R. Below are just a
few that, depending on your personal preferences and project needs,
may be better or worse than RStudio.

- R-app (comes with the R installation)
- Terminal/shell (R executable needs to be in your path)
- [Emacs + ESS](https://ess.r-project.org)
- [Atom](https://atom.io) + [r-exec](https://atom.io/packages/r-exec)
- [Jupyter](http://jupyter.org) + [IRkernel](https://irkernel.github.io)

## Miscellanea

- R script files end in `.r` or `.R`
- 1-indexed (indexes start at 1 instead of 0)
- Base commands usually written in C/C++ under the hood
- Can be run in batch mode from the terminal/command line
  - older: `R CMD BATCH`
  - newer: `Rscript`

# Notes
[^f1]: The "data scientist" as a person/title, like "big data," has
    probably become a little played out, but for lack of a better
    catch-all term, I think everyone knows what I mean.
[^f2]: For a little more history on R, particularly its success as an
    open source project, see [Fox
    (2009)](https://journal.r-project.org/archive/2009-2/RJournal_2009-2_Fox.pdf)
[^f3]: If you come to R knowing C/C++, Fortran, or Java, see
    [Rcpp](http://www.rcpp.org), [rFortran](http://www.rfortran.org),
    [rJava](https://www.rforge.net/rJava/index.html) for some cool
    interactivity.
[^f4]: Stata has some object-oriented features and R some procedural
    programming behaviors, so the assigned labels aren't perfect. They
    are mostly right, though.
[^f5]: Full disclosure: all questions I asked when learning R.
[^f6]: R also supports arrays, which can take on more than two dimensions.
[^f7]: For clarity, I'll call them packages when talking about what is
    downloaded and libraries when discussing what is loaded into
    memory. Since the names are the same, it's really a semantic
    difference.

```{r, echo = FALSE, purl = TRUE}

## =============================================================================
## END SCRIPT
################################################################################
```
