---
layout: module
title: Wrangling data I
date: 2018-01-01 00:00:03
category: module
links:
  script: wrangle_one.R
  data: els_plans.csv
output:
  md_document:
    variant: markdown_mmd
    preserve_yaml: true
---

```{r, include = FALSE, purl = FALSE}
source('knit_setup.R')
```
```{r, include = FALSE, purl = TRUE}
################################################################################
##
## <PROJ> R Workshop
## <FILE> wrangle_one.R 
## <INIT> 15 January 2018
## <AUTH> Benjamin Skinner (GitHub/Twitter: @btskinner)
##
################################################################################

## clear memory
rm(list = ls())

```
Being able to read, manipulate, and save data, that
is, [wrangle data](https://en.wikipedia.org/wiki/Data_wrangling), is a
key part of any analysis. In fact (as you're probably already aware),
building and cleaning data usually takes more time and lines of code
than the actual analysis.

In this module, I'm going to show you some data wrangling procedures,
using only base R functions. There's much to be
said for the tidyverse way of doing things (which we'll cover in the
next module), but I think it's still good to know how to use
core commands for those edge cases where tidyverse
functions don't quite work the way you want.

Data for this module come from the public release files of
the
[NCES Education Longitudinal Study of 2002](https://nces.ed.gov/surveys/els2002/). For
descriptions of the variables, see the <a href = '{{ site.baseurl
}}/data/#els_planscsv'>codebook</a>.

# Data wrangling with base R
```{r, echo = FALSE, purl = TRUE}
## -----------------------------------------------------------------------------
## Data wrangling with base R
## -----------------------------------------------------------------------------

```

First things first, let's read in the data. Base R can `load()` its
own data formats, `.rda` and `.RData`, as well as read flat
files like `.txt`, `.csv`, and `.tsv` files. (We'll discuss how to
read in data files from other languages later.) 

Since the data come in a CSV file, we could use the special command
`read.csv()`. The more generic function `read.table()` works just as
well, though, as long as we tell R that items in each row of our data
file are separated by a `,` using the `sep = ','` argument. By
default, R assumes that the data just begin, but since our file has
the variable names in the first row, we also need to use `header =
TRUE`. We won't talk about factors until later, but let's read in the
data keeping string values as character vectors as well.

```{r}
## read in the data, making sure that first line is read as column names
df <- read.table('../data/els_plans.csv', sep = ',', header = TRUE,
                 stringsAsFactors = FALSE)

```

## ...error!

If you tried to read the data and got an error that looked like this
```
Error in file(file, "rt") : cannot open the connection
In addition: Warning message:
In file(file, "rt") :
  cannot open file '../data/els_plans.csv': No such file or directory
```

then you have one of three issues:

1. You haven't downloaded the data (get it from link at top of page)
2. You don't have your directory/folder structure set up correctly (go
   back to the <a href="{{ site.baseurl
   }}/modules/#directory-structure">main module page</a> for information
   about how you should set things up)
3. You don't have R set to the correct working directory

To check your current working directory, use the `getwd()` function.

```{r, eval = FALSE, purl = TRUE}
## check current directory
getwd()

```

If you had a problem, it's possible that you aren't in the right
working directory. This means that R can't find your data because even
though it has directions on where to find it, it's starting in the
wrong spot so the directions are worthless!

If you know where your `scripts` directory/folder is, you can use the
`setwd()` function to change your working directory to the right one.

```{r}
## set the working directory, uncomment and change <path>/<to> as needed
## setwd('<path>/<to>/rworkshop/scripts')

```

You can correctly set the working directory by opening the script
using Rstudio's *Files* menu, which is in the bottom right-hand
box. Once you've found your script, use the *More* menu option "Set As
Working Directory" to set the `scripts` directory to the working
directory. Try reading the data again.

# Viewing data

Let's look at the first few rows and the variable names using the
`head()` function. RStudio makes it easy to see your data by using
its
[viewer](https://support.rstudio.com/hc/en-us/articles/205175388-Using-the-Data-Viewer),
(which really just calls the `View()` function around the data
object).

```{r}
## show the first few rows (or view in RStudio's view)
head(df)

```

Remember that we can also use `names()` to see just the variable names.

```{r}
## show the column names
names(df)

```

## Add variables
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## add variables
## ---------------------------

```

Add a column by giving it a name and assigning what you want. R will
repeat the values as necessary to fill the number of rows. You can
also use data from other columns. R will assign values row by row,
using the right-hand side values that align.


```{r}
## add simply column of ones
df$ones <- 1

## add sum of test scores (bynels2r + bynels2m)
df$sum_test <- (df$bynels2r + df$bynels2m)

## check names
names(df)

```

> #### Quick exercise
> Create a new column that is the average of the test scores.

## Drop variables
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## drop variables
## ---------------------------

```
Drop variables by assigning `NULL` to the column name.

```{r}
## drop follow up one panel weight
df$f1pnlwt <- NULL

## check names
names(df)

```

## Conditionally change values
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## conditionally change values
## ---------------------------

```
This can be tricky at first. To conditionally change or assign values,
you need to tell R where the conditions apply. There are a couple of
ways. 

The first way uses brackets, `[]`, after the variable name to set the
condition where the assignment is true. For version 1 below, the new
variable `female` is assigned a value of 1 in the rows where it is
`TRUE` that `bysex == 'female'`. In the rows that's `FALSE`, R will
assign `NA` since there's no information. We can back fill 0s in the
second line.

The other way is to use the `ifelse(test, yes, no)` function. Going
row by row, the `test` (`bysex == 'female'`) is performed. If `TRUE`,
the new variable gets a 1; if `FALSE`, it gets a 0.

```{r}
## make a numeric column that == 1 if bysex is female, 0 otherwise
## v.1
df$female_v1[df$bysex == 'female'] <- 1
df$female_v1[df$bysex != 'female'] <- 0

## v.2
df$female_v2 <- ifelse(df$bysex == 'female', 1, 0)

## the same?
identical(df$female_v1, df$female_v2)

```

> #### Quick exercise
> Create a new column called `ses_gender` that uses `byses1` for women
> and `byses2` for men. (HINT: if you use a condition, you need to use
> it on both sides of the arrow.)

## Filter
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## filter
## ---------------------------

```

You can also use brackets to conditionally drop rows, such as those
with missing values. 
```{r}
## assign as NA if < 0
df$bydob_p[df$bydob_p < 0] <- NA
nrow(df)

## drop if NA
df <- df[!is.na(df$bydob_p),]
nrow(df)

```

> #### Quick exercise
> The variable `byrace` also uses negative values to represent missing
> values. Reassign `NA`s to values that are less than zero. Next drop
> observations from the data set if they are missing `byrace`
> values. (HINT 1: Pay attention to your commas each time!) (HINT 2:
> Before dropping observations, save your data set object, `df` in
> another object, `df_hold`, just in case things don't go well the
> first time...)

## Order
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## order
## ---------------------------

```
Sort the data frame using the `order()` function as a condition.
```{r}
## show first few rows of student and base year math scores
df[1:10, c('stu_id','bydob_p')]

## since a data frame has two dims, notice the comma in the brackets
df <- df[order(df$bydob_p),]

## show again first few rows of ID and DOB
df[1:10, c('stu_id','bydob_p')]

```

> #### Quick exercise
> Sort by `bydob_p` again, but this time from youngest to oldest, that
> is, reverse order. Type `?order` to see the help file for the
> function and the argument you might need to do that.

## Aggregate
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## aggregate
## ---------------------------

```

To collapse the data, generating some summary statistic in the
process, use the `aggregate(x, by, FUN)`, where `x` is the data frame,
`by` is the grouping variable in a `list()`, and `FUN` is the function
that you want to use. These can be base R functions or one you create
yourself. Let's get the average math score within each school.

```{r}
## first, make test score values < 0 == NA
df$bynels2m[df$bynels2m < 0] <- NA

## create new data frame
sch_m <- aggregate(df$bynels2m, by = list(df$sch_id), FUN = mean, na.rm = T)

## show
head(sch_m)

```

> #### Quick exercise
> Get the average reading score for each school and save it to a
> different object. Don't forget to account for missing values.

## Merge
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## merge
## ---------------------------

```

Since you can have multiple data frames in memory (as objects) at the
same time in R, you may not find yourself merging data sets as often
you would in another language (like Stata, where you have to). That
said, it still needs to happen. Use the `merge()` function. Let's
merge the aggregated test score data back into the data set.

```{r}
## first fix names from aggregated data set
names(sch_m) <- c('sch_id', 'sch_bynels2m')

## merge on school ID variable
df <- merge(df, sch_m, by = 'sch_id')

## show
head(df)

```

> #### Quick exercise 
> Merge the average school level reading score data set you created to
> the full data set.


## Write
```{r, echo = FALSE, purl = TRUE}
## ---------------------------
## write
## ---------------------------

```

Finally we can write our new data set to disk. We can save it as an R
data file type, but since we may want to share with non-R users, we'll
save it as a csv file again.

```{r, eval = FALSE, purl = TRUE}
write.csv(df, '../data/els_plans_mod.csv', row.names = FALSE)
```

```{r, echo = FALSE, purl = TRUE}

## =============================================================================
## END SCRIPT
################################################################################
```
